---
title: "Project: Analysis on US Health Insurance Cost"
author: "Jenish, Suvid, Sakina"
output:
  pdf_document:
    fig_height: 3.2
    fig_width: 5
    latex_engine: xelatex
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
editor_options: 
  markdown: 
    wrap: 72
---

## Introduction

This report provides an exploratory data analysis (EDA) of the US health insurance dataset to understand its structure, identify trends, and prepare it for further analysis.

## Loading Required Libraries

```{r}
library(ggplot2)
library(dplyr)
library(summarytools)
library(gridExtra)
library(mplot)
```

## Loading the Dataset

```{r}
# Load dataset (replace 'health_insurance.csv' with your actual file path)
data <- read.csv("insurance.csv")

# Display the first few rows
head(data)
```

## Dataset Overview

### Summary Statistics

```{r}
summary(data)
```

### Checking Data Structure

```{r}
str(data)
```

### Checking for Missing Values

```{r}
colSums(is.na(data))
```

## Univariate Analysis

### Combined Plots

```{r}
plot_age <- ggplot(data, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "#1f77b4", color = "black") +
  labs(title = "Age Distribution", x = "Age", y = "Count")

plot_bmi <- ggplot(data, aes(x = bmi)) +
  geom_histogram(binwidth = 2, fill = "#ff7f0e", color = "black") +
  labs(title = "BMI Distribution", x = "BMI", y = "Count")

plot_charges <- ggplot(data, aes(x = charges)) +
  geom_histogram(binwidth = 2000, fill = "#2ca02c", color = "black") +
  labs(title = "Charges Distribution", x = "Charges", y = "Count")

grid.arrange(plot_age, plot_bmi, plot_charges, ncol = 3)
```

## Categorical Variables

### Combined Plots

```{r}
plot_gender <- ggplot(data, aes(x = sex)) +
  geom_bar(fill = "#9467bd") +
  labs(title = "Gender Distribution", x = "Gender", y = "Count")

plot_smoker <- ggplot(data, aes(x = smoker)) +
  geom_bar(fill = "#d62728") +
  labs(title = "Smoking Status", x = "Smoker", y = "Count")

plot_region <- ggplot(data, aes(x = region)) +
  geom_bar(fill = "#8c564b") +
  labs(title = "Region Distribution", x = "Region", y = "Count")

grid.arrange(plot_gender, plot_smoker, plot_region, ncol = 3)
```

## Bivariate Analysis

### Charges vs. Age

```{r}
ggplot(data, aes(x = age, y = charges)) +
  geom_point(alpha = 0.5, color = "#17becf") +
  geom_smooth(method = "lm", color = "#e377c2", se = FALSE) +
  labs(title = "Charges vs. Age", x = "Age", y = "Charges")
```

### Charges by Smoking Status

```{r}
ggplot(data, aes(x = smoker, y = charges, fill = smoker)) +
  geom_boxplot() +
  scale_fill_manual(values = c("#1f77b4", "#ff7f0e")) +
  labs(title = "Charges by Smoking Status", x = "Smoking Status", y = "Charges")
```

### Charges by Region

```{r}
ggplot(data, aes(x = region, y = charges, fill = region)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Charges by Region", x = "Region", y = "Charges")
```

### BMI vs. Charges

```{r}
ggplot(data, aes(x = bmi, y = charges)) +
  geom_point(alpha = 0.5, color = "#2ca02c") +
  geom_smooth(method = "lm", color = "#e377c2", se = FALSE) +
  labs(title = "BMI vs. Charges", x = "BMI", y = "Charges")
```

## Correlation Analysis

### Correlation Matrix

```{r}
# Correlation matrix for numerical variables
cor_matrix <- cor(data %>% select_if(is.numeric))
cor_matrix
```

## RQ1:

```{r}
library(ggplot2)
library(gridExtra)

# Plot 1: Charges vs Age with a fitted line
p1 <- ggplot(data, aes(x = age, y = charges)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) + # Add linear regression line
  labs(title = "Charges vs Age", x = "Age", y = "Charges") +
  theme_minimal()

# Plot 2: Charges vs BMI with a fitted line
p2 <- ggplot(data, aes(x = bmi, y = charges)) +
  geom_point(color = "green") +
  geom_smooth(method = "lm", color = "red", se = FALSE) + # Add linear regression line
  labs(title = "Charges vs BMI", x = "BMI", y = "Charges") +
  theme_minimal()

# Plot 3: Charges by Smoking Status (boxplot)
p3 <- ggplot(data, aes(x = smoker, y = charges, fill = smoker)) +
  geom_boxplot() +
  labs(title = "Charges by Smoking Status", x = "Smoking Status", y = "Charges") +
  theme_minimal()

# Plot 4: Charges by Region (boxplot)
p4 <- ggplot(data, aes(x = region, y = charges, fill = region)) +
  geom_boxplot() +
  labs(title = "Charges by Region", x = "Region", y = "Charges") +
  theme_minimal()

# Adjust plotting device size
options(repr.plot.width = 20, repr.plot.height = 15) # Adjust dimensions for better layout

# Arrange all plots in a grid
grid.arrange(p1, p2, p3, p4, ncol = 2, heights = c(1, 1))

```

```{r}
model <- lm(charges ~ age + smoker + bmi + bmi*smoker, data = data)
summary(model)

```

## RQ2:

```{r}
# Filter data for smokers and non-smokers
smokers <- data %>% filter(smoker == "yes")
non_smokers <- data %>% filter(smoker == "no")

# Create summary statistics for smokers and non-smokers
summary_stats <- data %>%
  group_by(smoker) %>%
  summarise(
    mean_charges = mean(charges),
    median_charges = median(charges),
    sd_charges = sd(charges),
    count = n()
  )

# Display the summary statistics
summary_stats

```


```{r}
# Perform a t-test to compare charges between smokers and non-smokers
t_test_result <- t.test(charges ~ smoker, data = data)

# View the results
t_test_result
```
## RQ3: Is there a significant difference in the insurance costs between male and female smokers?

### Hypotheses

-   **Null Hypothesis** ($H_0$): The mean insurance costs for male and female smokers are equal.
-   **Alternative Hypothesis** (\$H_1\$):The mean insurance costs for male and female smokers are not equal.

To answer this question, we will use a two-sample t-test.

```{r}
# Filter data for smokers only
smokers <- data %>%
  filter(smoker == "yes")

# Create separate groups for male and female smokers
male_smokers <- smokers %>% filter(sex == "male") %>% pull(charges)
female_smokers <- smokers %>% filter(sex == "female") %>% pull(charges)

# Display summary statistics
summary_stats <- smokers %>%
  group_by(sex) %>%
  summarise(
    mean_charges = mean(charges),
    median_charges = median(charges),
    sd_charges = sd(charges),
    count = n()
  )

```

```{r}

# Display the summary table
knitr::kable(summary_stats, caption = "Summary Statistics for Male and Female Smokers")
```

### Boxplot to visualize the distribution of insurance charges

```{r}

ggplot(smokers, aes(x = sex, y = charges, fill = sex)) +
  geom_boxplot() +
  labs(
    title = "Insurance Costs for Male and Female Smokers",
    x = "Gender",
    y = "Insurance Charges"
  ) +
  theme_minimal()

```

### Perform a two-sample t-test

```{r}
t_test_result <- t.test(male_smokers, female_smokers, var.equal = TRUE)

# Display the test results
t_test_result

```

#### Summary
Since the p-value (0.09448) is greater than 0.05 and the confidence interval includes zero, we fail to reject the null hypothesis. This suggests that there is no statistically significant difference in the mean insurance charges between male smokers and female smokers at the 5% significance level. However, there might still be a practical difference in the mean charges, as the confidence interval is quite wide, indicating variability in the data.

## RQ4: Is there a significant difference in charges for individuals in different regions?

Is there a significant difference in insurance charges among individuals living in different regions (northeast, northwest, southeast, southwest)?

### Hypotheses

-   **Null Hypothesis** ($H_0$): The mean insurance charges are equal across all regions.
-   **Alternative Hypothesis** ($H_1$): At least one region has a significantly different mean insurance charge.

We will use a one-way ANOVA test to answer this research question.

------------------------------------------------------------------------

```{r setup, include=TRUE}
# Summarize charges by region
summary_by_region <- data %>%
  group_by(region) %>%
  summarise(
    mean_charges = mean(charges),
    median_charges = median(charges),
    sd_charges = sd(charges),
    count = n()
  )
```

```{r}
# Display the summary table
knitr::kable(summary_stats, caption = "Summary Statistics for Male and Female Smokers")
```

### **Boxplot to visualize the distribution of charges by region**

```{r}
ggplot(data, aes(x = region, y = charges, fill = region)) +
  geom_boxplot() +
  labs(
    title = "Insurance Charges by Region",
    x = "Region",
    y = "Charges"
  ) +
  theme_minimal()
```

### Perform one-way ANOVA

```{r}
anova_result <- aov(charges ~ region, data = data)

# Display the ANOVA table
summary(anova_result)
```

#### Summary
The results of the one-way ANOVA test indicate a statistically significant difference in mean insurance charges across the four regions (northeast, northwest, southeast, southwest). The F-statistic is 2.97, with a p-value of 0.0309, which is below the significance level of 0.05. Thus, we reject the null hypothesis that the mean insurance charges are the same across all regions. This finding suggests that regional differences in insurance charges exist. However, the ANOVA test does not specify which regions differ from each other. To determine the specific pairwise differences, we perform a post hoc analysis such as Tukey’s Honest Significant Difference (HSD) test.

### Tukey's Honest Significant Difference (HSD) test

```{r}
post_hoc <- TukeyHSD(anova_result)

# Display the results
post_hoc
```

```{r}
# Convert Tukey HSD results to a data frame
tukey_df <- as.data.frame(post_hoc$region)
tukey_df$Comparison <- rownames(tukey_df)

# Plot the results
ggplot(tukey_df, aes(x = Comparison, y = diff)) +
  geom_point(size = 3, color = "blue") +
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.2, color = "red") +
  labs(
    title = "Tukey HSD Results",
    x = "Region Comparison",
    y = "Difference in Means"
  ) +
  theme_minimal() +
  coord_flip()


```

#### Summary

From this plot, we can observe which region pairs show significant differences. For region pairs where the confidence interval does not include zero (southwest-southeast), we conclude that the insurance charges differ significantly. However, rest of the regions there is insufficient evidence to suggest a significant difference in insurance charges.
